oREV: an Item Response Theory based open receptive vocabulary task for 3 to 8-year-old children

### 2022-08-13 paper/manuscript.Rmd
work in Natalias comments

@@

abstract: |
  Individual differences in early language abilities are an important predictor of later life outcomes. High-quality, easy-access measures of language abilities are rare, especially in the preschool years. The present study describes the construction of a new receptive vocabulary task for children between 3 and 8 years of age. The task was implemented as a browser-based web application, allowing for in-person as well as remote data collection via the internet. Based on data from N = 581 German-speaking children, we estimated the psychometric properties of each item in a larger initial item pool via Item Response Modeling. We then applied an automated item selection procedure to select an optimal subset of items based on item difficulty and discrimination. The so-constructed task has 20 items and correlates with the full task (52 items) at a rate of .97. The construction, implementation and item selection process we described here makes it easy to extend the task or adapt it to different languages. All materials and code are freely accessible to interested researchers.

@@

Individual differences in language abilities are early emerging, stable across development and predictive of a wide range of psychological outcome variables including cognitive abilities, academic achievement or mental health [@bornstein2018stability; @schoon2010children; @morgan201524; @marchman2008speed; @walker1994prediction]. From a methodological perspective, high-quality, easy-access measures of language abilities are therefore central to both basic and applied research on individual differences in language abilities. Ideally, such measures should also be comparable across languages in order to study which developmental processes are language-specific and which are shared more widely. Developing such measures is very time and resource intensive and, as a consequence, few exist. In this paper, we describe the construction of a new receptive vocabulary task for German-speaking children. Its psychometric grounding in Item Response Theory makes the measure robust and efficient. Its web-based design and implementation makes the measure easy to adapt and administer in different settings (in-person or remote) and thereby facilitates the scaling of data collection.

Language has many facets and aspects that can be focused on when assessing individual differences between children. One particular productive approach has been the study of children's vocabulary skills, that is, their knowledge of word-object mappings. This skill can be easily assessed, for example by asking children to name an object (productions) or pick out an object that matches a word they just heard (comprehension). Children with larger vocabularies are taken to have advanced language skills more broadly. This assumption seems to be justified in light of strong correlations between vocabulary size and other language measures such as grammatical skills [e.g., @moyle2007longitudinal; @hoff2018explains].

@@

From 2 years onward, children are usually tested directly. Vocabulary assessment is often part of standardized tests of cognitive abilities [e.g. @wechsler1949wechsler; @gershon2013iv; @bayley2006bayley]. In addition, a range of dedicated forms exist [e.g., @dunn1965peabody; @dunn1997british; @golinkoff2017user]. Yet, from the perspective of a researcher, these existing measures are often problematic for several reasons. Because they are standardized and normed instruments using them ensues substantial licensing costs. For the same reasons, the corresponding materials are not openly available which makes it difficult to expand or adapt them to different languages. Most measures also rely on in-person, paper-pencil testing which makes large-scale data collection inefficient and extremely costly. Whenever more portable, computerized versions exist, they come with additional costs. As a consequence, nothing comparable to the collaborative research infrastructure built around the CDI exists for vocabulary measures for older children.

The development of so-called Cross-linguistic Lexical Tasks (ClTs; @haman2015designing) constitutes a promising framework that might help to overcome these issues. CLTs are simple picture-choice and picture-naming tasks aimed at assessing comprehension and production of nouns and verbs. In a collaborative effort involving more than 25 institutions, versions for dozens of different languages have been developed following the same guiding principles [see @haman2015designing; @haman2017noun; @armon2015assessing]. In addition to cross-linguistic studies with mono-lingual children, this procedure makes CLTs ideally suited to assess multi-lingual children. The tasks and the materials are not commercially licensed and can thus be freely used for research purposes.

Despite these many positive characteristics, CLTs are limited in two important ways. First, they show ceiling effects for receptive vocabulary skills of mono-lingual children older than 3. This greatly limits their usefulness in research across the preschool years. Second, and maybe more important, CLTs have been developed following clear linguistic guidelines -- but *without* a psychometric framework^[The same applies to most other vocabulary measures used in developmental research.]. As a consequence, it is unclear how the different items relate to the underlying construct (e.g., vocabulary skills). We do not know which items discriminate between different levels of skill and are therefore particularly diagnostic e.g., at different ages. Items could also be biased and show differential measurement properties in relevant subgroups (e.g. girls and boys). In addition, some items might be simply redundant in that they measure the underlying construct in the same way. Such characteristics could make the task unnecessarily long. Modern psychometric approaches like Item Response Theory (IRT) [@lord2012applications; @kubinger2006psychologische] assess the relation between each individual item and the underlying -- latent -- construct one seeks to measure. This focus allows for evaluating the quality and usefulness of each item and thereby provides a solid psychometric basis for constructing efficient and high-quality tasks. In combination with a computerized implementation, IRT allows for adaptive testing during which participants are selectively presented with highly informative items given their (constantly updated) estimated level of ability. However, IRT-based task construction requires a higher initial investment: it takes a large item pool and large sample sizes to estimate the item parameters that guide the selection of the best items.

@@

The initial item pool consisted of 32 items taken from the German CLT [@haman2015designing; @haman2017noun] and 20 new items. The addition of new items was necessary due to ceiling effects for monolingual 3-year-olds in the previous version. New items were generated in line with the construction of the original CLT in a stepwise process. Each item consists of a target word and three distractors. To select target words, we first compiled a list of age-of-acquisition ratings for 3,928 German words from various sources [@luniewska2019age; @schroder2012german; @birchenough2017rated]. From this list, we selected 20 words based on the following criteria: words should refer to concepts that could easily and unambiguously be depicted in a drawing, age-of-acquisition ratings should be spread equally between six and ten years of age, and words should have comparable complexity indices [see @haman2017noun]. The so-selected 20 words served as additional target words in the item pool (total of 52 items). For each target word, we selected three distractors. The first distractor was unrelated to the target word but was chosen to have a comparable rated age-of-acquisition. The second distractor was semantically related to the target word (e.g. ruin -- fortress; elk -- mammoth). The third distractor was phonetically similar to the target, that is ... (e.g. Gazelle [eng.: gazelle] -- Libelle [eng.: dragonfly]). The full list of targets and distractors can be found in the associate online repository. Finally, an artist (same as for the original CLT items) drew pictures representing all target and distractor words. This procedure ensured that the original CLT and the newly generated items formed a homogeneous item pool.

@@

The task was programmed in `JavaScript` and `HTML` and presented as a website which could be opened in any modern web browser. In addition to participants' responses, we recorded webcam videos^[Due to access rights issues, webcam recording was not possible when participants used iOS devices.]. Both files were sent to a server after the study was finished. The task started with several instruction pages that explained to parents the task and how they should assists their child if needed. The task can be accessed via the following link: https://ccp-odc.eva.mpg.de/clt-extended/.

@@

```{r}
data <- read_csv("../data/clt_clean_data.csv")

@@

The automated item selection process we implemented critically leveraged the strengths of IRT modelling. For each item in the pool we estimated its difficulty and discrimination. We defined the object function we optimized via the simulated annealing process so that it would yield a subset in which items would a) be equally spread out across the latent ability so that the task measured equally well at different skill levels and b) have maximal discrimination so that the items differentiate well between individuals having similar skill levels. In addition, we prioritized items with more precise difficulty estimates (i.e. narrower CrIs). This procedure presents a principled way of constructing a task with good psychometric properties which can easily be applied to any new set of items or versions of the task in different languages. However, this approach does not make the careful, principle-based construction of the initial item pool superfluous; it only selects the best of the available items. Linguistic and psychometric considerations thus need to go hand in hand during task construction.

@@

We have described the construction of a new receptive vocabulary measure for German-speaking children between 3 and 8 years-of-age. The datastes and the analysis code for item selection are freely available in the associated online repository (https://github.com/ccp-eva/vocab). An online version of the task is available at the following website: . The implementation architecture (`JavaScript` and `HTML` code) and the materials can be accessed in the following repository: https://github.com/ccp-eva/CLT-extended. These resources allow interested researchers to use, extend and adapt the task.

>>>>

  - name          : "Büsra Delikaya"
    affiliation   : "2"
    role:
      - Writing - Review & Editing

@@

abstract: |
  Individual differences in early language abilities are an important predictor of later life outcomes. High-quality, easy-access measures of language abilities are rare, especially in the preschool years. The present study describes the construction of a new receptive vocabulary task for children between 3 and 8 years of age. The task was implemented as a browser-based web application, allowing for in-person as well as remote data collection via the internet. Based on data from N = 581 German-speaking children, we estimated the psychometric properties of each item in a larger initial item pool via Item Response Modeling. We then applied an automated item selection procedure to select an optimal subset of items based on item difficulty and discrimination. The so-constructed task has 20 items and correlates with the full task (52 items) at a rate of .97. The construction, implementation and item selection process we described here makes it easy to extend the task or adapt it to different languages. All materials and code are freely accessible to interested researchers. The tasked can be used via the following website: https://ccp-odc.eva.mpg.de/orev-demo/.

@@

Individual differences in language abilities are early emerging, stable across development, and predictive of a wide range of psychological outcome variables including cognitive abilities, academic achievement, or mental health [@bornstein2018stability; @schoon2010children; @morgan201524; @marchman2008speed; @walker1994prediction]. From a methodological perspective, high-quality, easy-access measures of language abilities are therefore central to both basic and applied research on individual differences in language abilities. Ideally, such measures should also be comparable across languages in order to study which developmental processes are language-specific and which are shared more widely. Developing such measures is very time and resource intensive and, as a consequence, few exist. In this paper, we describe the construction of a new receptive vocabulary task for German-speaking children. Its psychometric grounding in Item Response Theory makes the measure robust and efficient. Its web-based design and implementation make the measure easy to adapt and administer in different settings (in-person or remote) and thereby facilitate the scaling of data collection.

Language has many facets and aspects that can be focused on when assessing individual differences between children. One particular productive approach has been the study of children’s vocabulary skills, that is, their knowledge of word-object mappings. This skill can be most effectively assessed, for example by asking children to name an object (production) or pick out an object that matches a word they just heard (comprehension). Children with larger vocabularies are taken to have advanced language skills more broadly. This assumption seems to be justified in light of strong correlations between vocabulary size and other language measures such as grammatical [e.g., @moyle2007longitudinal; @hoff2018explains] or narrative skills [@bohnacker2021storytelling; @fiani2021macrostructure; @lindgren2022age; @tsimpli2016narrative]. Vocabulary skills have also been used as an indicator of developmental language disorders more broadly [@spaulding2013investigating]. Finally, many of the predictive relations found for early language skills mentioned above are based on vocabulary measures [@bleses2016early; @golinkoff2019language; @pace2017identifying; @pace2019measuring]. This set of findings underlines the importance of high-quality vocabulary measures.

@@

However, the CDI best suited for children in the first two years of life. From 2 years onward, children are usually tested directly. Vocabulary assessment is often part of standardized tests of cognitive abilities [e.g. @wechsler1949wechsler; @gershon2013iv; @bayley2006bayley]. In addition, a range of dedicated forms exist for English [e.g., @dunn1965peabody; @dunn1997british; @golinkoff2017user], German [@kiese2005awst; @kauschke2002patholinguistische; @lenhard2015peabody; @gluck2011wortschatz] and other languages.

Yet, from the perspective of a researcher, these existing measures are often problematic for several reasons. Because they are standardized and normed instruments using them ensues substantial licensing costs. For the same reasons, the corresponding materials are not openly available which makes it difficult to expand or adapt them to different languages. Most measures also rely on in-person, paper-pencil testing which makes large-scale data collection inefficient and extremely costly. Whenever more portable, computerized versions exist, they come with additional costs. As a consequence, nothing comparable to the collaborative research infrastructure built around the CDI exists for vocabulary measures for older children.

The development of so-called Cross-linguistic Lexical Tasks (ClTs; @haman2015designing) constitutes a promising framework that might help to overcome these issues. CLTs are simple picture-choice and picture-naming tasks aimed at assessing comprehension and production of nouns and verbs. In a collaborative effort involving more than 25 institutions, versions for dozens of different languages have been developed following the same guiding principles [@haman2015designing; @haman2017noun; @armon2015assessing]. In addition to cross-linguistic studies with mono-lingual children, this procedure makes CLTs ideally suited to assess multi-lingual children. The tasks and the materials are not commercially licensed and can thus be freely used for research purposes.

Despite these many positive characteristics, CLTs are limited in two important ways. Despite these many positive characteristics, CLTs are limited in two important ways. First, they were designed for children between 3 and 5 years and consequently they show ceiling effects for older children in this age range [@haman2017noun]. This greatly limits their usefulness in research across the preschool years. Second, and maybe more important, CLTs have been developed following clear linguistic guidelines -- but *without* a strict psychometric framework^[The same applies to most other vocabulary measures used in developmental research.]. As a consequence, it is unclear how the different items relate to the underlying construct (e.g., vocabulary skills). We do not know which items discriminate between different levels of skill and are therefore particularly diagnostic e.g., at different ages. Items could also be biased and show differential measurement properties in relevant subgroups (e.g. girls and boys). In addition, some items might be simply redundant in that they measure the underlying construct in the same way. Such characteristics could make the task unnecessarily long. Modern psychometric approaches like Item Response Theory (IRT) [@lord2012applications; @kubinger2006psychologische] assess the relation between each individual item and the underlying -- latent -- construct one seeks to measure. This focus allows for evaluating the quality and usefulness of each item and thereby provides a solid psychometric basis for constructing efficient and high-quality tasks. In combination with a computerized implementation, IRT allows for adaptive testing during which participants are selectively presented with highly informative items given their (constantly updated) estimated level of ability. However, IRT-based task construction requires a higher initial investment: it takes a large item pool and large sample sizes to estimate the item parameters that guide the selection of the best items.

@@

The initial item pool consisted of 32 items taken from the German CLT [@haman2015designing; @haman2017noun] and 20 new items. The addition of new items was necessary due to ceiling effects for monolingual 5-year-olds in the previous version. New items were generated in line with the construction of the original CLT in a stepwise process. Each item consists of a target word and three distractors. To select target words, we first compiled a list of age-of-acquisition ratings for 3,928 German words from various sources [@luniewska2019age; @schroder2012german; @birchenough2017rated]. From this list, we selected 20 words based on the following criteria: words should refer to concepts that could easily and unambiguously be depicted in a drawing, age-of-acquisition ratings should be spread equally between six and ten years of age. We also computed complexity indices for each word [see @haman2017noun]. This metric, however, did not reflect a dimension that was relevant for item selection.

The so-selected 20 words served as additional target words in the item pool (total of 52 items). For each target word, we selected three distractors. The first distractor was unrelated to the target word but was chosen to have a comparable rated age-of-acquisition. The second distractor was semantically related to the target word (e.g. ruin -- fortress; elk -- mammoth). The third distractor was phonetically similar to the target, that is, for example, initial part was substituted, and the rest of the word was kept similar (e.g. Gazelle [eng.: gazelle] -- Libelle [eng.: dragonfly]). The full list of targets and distractors can be found in the associate online repository. Finally, an artist (same as for the original CLT items) drew pictures representing all target and distractor words. This procedure ensured that the original CLT and the newly generated items formed a homogeneous item pool.

@@

The task was programmed in `JavaScript` and `HTML` and presented as a website which could be opened in any modern web browser. In addition to participants' responses, we recorded webcam videos^[Due to access rights issues, webcam recording was not possible when participants used iOS devices.]. Both files were sent to a server after the study was finished. The task started with several instruction pages that explained to parents the task and how they should assist their child if needed. The task (after item selection) can be accessed via the following link: https://ccp-odc.eva.mpg.de/orev-demo/.

@@

```{r}
data <- read_csv("../data/clean_data.csv")

@@

The automated item selection process we implemented critically leveraged the strengths of IRT modelling. For each item in the pool we estimated its difficulty and discrimination. We defined the object function we optimized via the simulated annealing process so that it would yield a subset in which items would a) be equally spread out across the latent ability so that the task measured equally well at different skill levels and b) have maximal discrimination so that the items differentiate well between individuals having similar skill levels. In addition, we prioritized items with more precise difficulty estimates (i.e. narrower CrIs).

This procedure presents a principled way of constructing a task with good psychometric properties which can easily be applied to any new set of items or versions of the task in different languages. However, this approach does not make the careful, principle-based construction of the initial item pool superfluous; it only selects the best of the available items. Linguistic and psychometric considerations thus need to go hand in hand during task construction. For example, while nouns are more similar across languages, verbs are more language-specific and might have different representations or even be absent as a single word, e.g. verb "wandern" (German for hiking) can be expressed only by an analytical construction in Slavic languages. Furthermore, bilingual and monolingual lexicons might vary and background factors, such as age, length of exposure or the onset of second language acquisition should be considered. Finally, morphosyntactic properties of verb grammar, such as perfective or imperfective aspect, should be considered.

@@

We have described the construction of a new receptive vocabulary measure for German-speaking children between 3 and 8 years-of-age. The datastes and the analysis code for item selection are freely available in the associated online repository (https://github.com/ccp-eva/vocab). An online version of the task is available at the following website: https://ccp-odc.eva.mpg.de/orev-demo/. The implementation architecture (`JavaScript` and `HTML` code) and the materials can be accessed in the following repository: https://github.com/ccp-eva/orev-demo. These resources allow interested researchers to use, extend and adapt the task.
